---
title: "Project 1 - Exploratory Data Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Data Import and Setup

```{r data-import}
# EDA - Import RDS file as dataframe
df <- readRDS("dat.rds")

# Load required packages
if (!require(dplyr)) install.packages("dplyr")
library(dplyr)
if (!require(ggplot2)) install.packages("ggplot2")
library(ggplot2)
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(stringr); library(ggplot2); library(forcats)
})

```

## Initial Data Structure

```{r data-structure}
# Display initial structure
cat("Original dataset dimensions:", dim(df), "\n")
print(paste("Total columns:", ncol(df)))

# Identify metadata columns vs measurement columns
metadata_cols <- grep("^Metadata_", colnames(df), value = TRUE)
measurement_cols <- setdiff(colnames(df), metadata_cols)

cat("Metadata columns:", length(metadata_cols), "\n")
cat("Measurement columns:", length(measurement_cols), "\n")
```

*Insert analysis here*

## Recover the data structure

```{r missing-data-analysis}
  plates_per_site <- 7
  reps_per_site   <- 4
  rep_levels      <- paste0("R", 1:reps_per_site)
  
  site_to_plot    <- "FMP" 
  
  controls <- c("DMSO", "Nocodazole", "Tetrandrine")
  control_color_map <- c("DMSO"="green","Nocodazole"="purple","Tetrandrine"="red",
                         "Other"="grey80","Missing"="black")
  missing_fill_color <- "grey30"
  tile_border_width  <- 0.25
  
  # =========================
  # Checks
  # =========================
  stopifnot(exists("df"))
  req_cols <- c("Metadata_Site","Metadata_Row","Metadata_Col",
                "Metadata_Object_Count","Metadata_Compound_type")
  miss <- setdiff(req_cols, names(df))
  if (length(miss)) stop("Missing cols: ", paste(miss, collapse=", "))
  
  plate_col <- c("Metadata_Unique_plate","Metadata_Plate_num","Metadata_Plate",
                 "Plate","PlateID","Barcode","Metadata_Barcode")
  plate_col <- plate_col[plate_col %in% names(df)][1]
  if (is.na(plate_col)) stop("Set `plate_col` to your plate ID column.")
  
  # =========================
  # Parse Plate & Replicate from plate string
  # =========================
  df_parsed <- df %>%
    mutate(
      Metadata_Row = as.character(Metadata_Row),
      Metadata_Col = suppressWarnings(as.numeric(Metadata_Col)),
      PlateRaw     = as.character(.data[[plate_col]]),
      .rep_digit   = str_match(PlateRaw, "(?i)(?:^|[_-])R\\s*([1-4])$")[,2],
      Replicate    = ifelse(!is.na(.rep_digit), paste0("R", str_trim(.rep_digit)), NA_character_),
      Plate        = ifelse(!is.na(Replicate),
                            str_replace(PlateRaw, "(?i)[_-]R\\s*[1-4]$", ""),
                            PlateRaw)
    ) %>%
    select(-.rep_digit) %>%
    filter(!is.na(Metadata_Col), Metadata_Col>=1, Metadata_Col<=24,
           Metadata_Row %in% LETTERS[1:16])
  
  to_well_coords <- function(d) d %>% mutate(Row_Num = match(Metadata_Row, LETTERS[1:16]),
                                             Col_Num = Metadata_Col)
  
  # Plate order helper (within THIS site)
  plate_order_for_site <- function(site_name) {
    p <- df_parsed %>% filter(Metadata_Site==site_name) %>% distinct(Plate) %>% pull(Plate)
    if (!length(p)) return(character(0))
    num <- suppressWarnings(as.numeric(str_extract(p, "\\d+")))
    p[order(is.na(num), num, p)]
  }
  
  # =========================
  # Build complete **4 × 7** layout for ONE site
  # =========================
  build_site_grid <- function(site_name) {
    wells <- expand_grid(Metadata_Row=LETTERS[1:16], Metadata_Col=1:24)
    
    # pick exactly 7 plates (pad with placeholders if fewer exist)
    p_order <- plate_order_for_site(site_name)
    sel_plates <- head(p_order, plates_per_site)
    if (length(sel_plates) < plates_per_site) {
      sel_plates <- c(sel_plates,
                      paste0("MISSING_PLATE_", seq_len(plates_per_site - length(sel_plates))))
    }
    
    # skeleton: 7 plates × 4 reps × wells
    skel <- expand_grid(
      Metadata_Site = site_name,
      Plate         = sel_plates,
      Replicate     = rep_levels
    ) %>%
      crossing(wells) %>%
      to_well_coords() %>%
      mutate(
        PlateColIdx = match(Plate, sel_plates),          # columns = plates
        RepRowIdx   = match(Replicate, rep_levels),      # rows    = replicates
        PlateFacet  = factor(sprintf("Plate %d\n%s", PlateColIdx, Plate),
                             levels = sprintf("Plate %d\n%s", 1:plates_per_site, sel_plates)),
        RepFacet    = factor(sprintf("Rep %d\n%s", RepRowIdx, Replicate),
                             levels = sprintf("Rep %d\n%s", 1:reps_per_site, rep_levels))
      )
    
    # actual averages and dominant control (filtered to this site)
    dat <- df_parsed %>%
      filter(Metadata_Site==site_name) %>%
      group_by(Metadata_Site, Plate, Replicate, Metadata_Row, Metadata_Col) %>%
      summarise(avg_cell_count=mean(Metadata_Object_Count, na.rm=TRUE), .groups="drop")
    
    ctrl <- df_parsed %>%
      filter(Metadata_Site==site_name) %>%
      group_by(Metadata_Site, Plate, Replicate, Metadata_Row, Metadata_Col, Metadata_Compound_type) %>%
      summarise(n=dplyr::n(), .groups="drop") %>%
      group_by(Metadata_Site, Plate, Replicate, Metadata_Row, Metadata_Col) %>%
      slice_max(order_by=n, n=1, with_ties=FALSE) %>%
      ungroup() %>%
      mutate(control_label = ifelse(Metadata_Compound_type %in% controls,
                                    Metadata_Compound_type, "Other")) %>%
      select(Metadata_Site, Plate, Replicate, Metadata_Row, Metadata_Col, control_label)
    
    merged <- skel %>%
      left_join(dat,  by=c("Metadata_Site","Plate","Replicate","Metadata_Row","Metadata_Col")) %>%
      left_join(ctrl, by=c("Metadata_Site","Plate","Replicate","Metadata_Row","Metadata_Col"))
    
    if (!"control_label" %in% names(merged)) merged$control_label <- NA_character_
    
    merged %>%
      mutate(
        is_missing    = is.na(avg_cell_count),
        control_label = ifelse(is_missing, "Missing", control_label)
        # keep avg_cell_count = NA so fill uses `missing_fill_color`
      )
  }
  
  site_df <- build_site_grid(site_to_plot)
  

```

*Insert analysis here*

## Missing Data Count

```{r missing-data-by-type}
# =========================
# Missingness — 4 separate heat maps (one per site)
# =========================
# Prereqs assumed from your script:
# - df_parsed (from your parsing block)
# - build_site_grid(), rep_levels, plates_per_site, tile_border_width
# - libraries: dplyr, ggplot2, tidyr, stringr (already in your code)

# Pick exactly 4 sites (alphabetical); change this vector if you prefer specific ones
sites_to_plot <- sort(unique(df_parsed$Metadata_Site))[1:4]
if (length(sites_to_plot) < 4) {
  warning("Fewer than 4 sites found; plotting whatever is available.")
}

# build missing/observed status df for a site
site_missing_df <- function(site_name) {
  build_site_grid(site_name) %>%
    dplyr::mutate(
      data_status = dplyr::if_else(is_missing, "Missing", "Observed"),
      data_status = factor(data_status, levels = c("Observed", "Missing"))
    )
}


# ---- 1) Per-site numeric summary (for the 4 sites) ----
missing_summary_4 <- lapply(sites_to_plot, site_missing_df) |>
  dplyr::bind_rows() |>
  dplyr::group_by(Metadata_Site) |>
  dplyr::summarise(
    total_wells    = dplyr::n(),
    missing_wells  = sum(data_status == "Missing"),
    observed_wells = sum(data_status == "Observed"),
    missing_pct    = round(100 * mean(data_status == "Missing"), 2),
    .groups = "drop"
  ) |>
  dplyr::arrange(desc(missing_pct))

cat("\n=== Missingness Summary (Selected 4 Sites) ===\n")
print(missing_summary_4)



```

*Insert analysis here*

## Missing Data Patterns

```{r missing-data-patterns}

# make the binary RAW-style missingness heat map for a site
make_missing_plot <- function(site_name) {
  sdf <- site_missing_df(site_name)
  ggplot(sdf, aes(x = Col_Num, y = Row_Num)) +
    geom_tile(aes(fill = data_status),
              color = "grey30", linewidth = tile_border_width) +
    facet_grid(RepFacet ~ PlateFacet, switch = "y", drop = FALSE) +  # 4 × 7 layout
    scale_fill_manual(
      values = c("Observed" = "white", "Missing" = "red"),
      name   = "Data Status"
    ) +
    scale_x_continuous(breaks = seq(1, 24, 4), expand = c(0, 0)) +
    scale_y_continuous(
      breaks = seq(1, 16, 2),
      labels = LETTERS[seq(1, 16, 2)],
      trans  = "reverse", expand = c(0, 0)
    ) +
    labs(
      title    = paste0("RAW Missingness — Site: ", site_name, " (4 Replicates × 7 Plates)"),
      subtitle = "Red = Missing, White = Observed. Layout identical to RAW heat map.",
      x = "Column", y = "Row"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      panel.grid = element_blank(),
      strip.text = element_text(face = "bold")
    ) +
    coord_fixed()
}

# ---- 2) Four separate heat maps (print one by one) ----
plots_4 <- lapply(sites_to_plot, make_missing_plot)
for (p in plots_4) print(p)
```

*Insert analysis here*



## Compound Cell Count Analysis

```{r compound-analysis-setup}
# COMPOUND CELL COUNT RANKING
cat("\n=== COMPOUND RANKING BY CELL COUNT ===\n")

# Use Metadata_EOS as compound identifier
compound_col <- "Metadata_EOS"
```

```{r compound-counts}
if(compound_col %in% colnames(df) && "Metadata_Object_Count" %in% colnames(df)) {
  # Count cells per compound using Metadata_Object_Count
  compound_counts <- df %>%
    group_by(!!sym(compound_col)) %>%
    summarise(
      cell_count = mean(Metadata_Object_Count, na.rm = TRUE),
      cell_count_sd = sd(Metadata_Object_Count, na.rm = TRUE),
      cell_count_min = min(Metadata_Object_Count, na.rm = TRUE),
      cell_count_max = max(Metadata_Object_Count, na.rm = TRUE),
      total_observations = n(),
      .groups = 'drop'
    ) %>%
    arrange(cell_count)
  
  cat("\nCompound cell count summary:\n")
  cat("Total unique compounds:", nrow(compound_counts), "\n")
  cat("Min cells per compound:", min(compound_counts$cell_count), "\n")
  cat("Max cells per compound:", max(compound_counts$cell_count), "\n")
  cat("Mean cells per compound:", round(mean(compound_counts$cell_count), 2), "\n")
  cat("Median cells per compound:", median(compound_counts$cell_count), "\n")
}
```

*Insert analysis here*

## Compound Type Analysis

```{r compound-type-analysis}
# Create summary by compound type and store reference lines
compound_type_summary <- NULL
dmso_avg <- NA
nocodazole_avg <- NA  
tetrandrine_avg <- NA
dmso_sd <- NA
nocodazole_sd <- NA
tetrandrine_sd <- NA

if("Metadata_Compound_type" %in% colnames(df)) {
  cat("\n=== COMPOUND TYPE ANALYSIS ===\n")
  
  compound_type_summary <- df %>%
    group_by(Metadata_Compound_type) %>%
    summarise(
      observations = n(),
      min_cell_count = min(Metadata_Object_Count, na.rm = TRUE),
      max_cell_count = max(Metadata_Object_Count, na.rm = TRUE),
      avg_cell_count = round(mean(Metadata_Object_Count, na.rm = TRUE), 2),
      sd_cell_count = round(sd(Metadata_Object_Count, na.rm = TRUE), 2),
      .groups = 'drop'
    ) %>%
    arrange(desc(observations))
  
  cat("Summary by compound type:\n")
  print(compound_type_summary)
  
  # Extract values for vertical lines
  if("DMSO" %in% compound_type_summary$Metadata_Compound_type) {
    dmso_avg <- compound_type_summary$avg_cell_count[compound_type_summary$Metadata_Compound_type == "DMSO"]
    dmso_sd <- compound_type_summary$sd_cell_count[compound_type_summary$Metadata_Compound_type == "DMSO"]
  }
  if("Nocodazole" %in% compound_type_summary$Metadata_Compound_type) {
    nocodazole_avg <- compound_type_summary$avg_cell_count[compound_type_summary$Metadata_Compound_type == "Nocodazole"]
    nocodazole_sd <- compound_type_summary$sd_cell_count[compound_type_summary$Metadata_Compound_type == "Nocodazole"]
  }
  if("Tetrandrine" %in% compound_type_summary$Metadata_Compound_type) {
    tetrandrine_avg <- compound_type_summary$avg_cell_count[compound_type_summary$Metadata_Compound_type == "Tetrandrine"]
    tetrandrine_sd <- compound_type_summary$sd_cell_count[compound_type_summary$Metadata_Compound_type == "Tetrandrine"]
  }
}
```

*Insert analysis here*

## Top Compounds with Least Cells

```{r least-cells-compounds}
# Show compounds with least cells
cat("\nTop 20 compounds with LEAST cells:\n")
print(head(compound_counts, 20))
```

*Insert analysis here*

## Cell Count Distribution Histogram

```{r cell-count-histogram}
# Create histogram of cell count distribution
cat("\nCreating histogram of compound cell count distribution...\n")

histogram_plot <- ggplot(compound_counts, aes(x = cell_count)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(
    title = "Distribution of Cell Counts per Compound",
    x = "Average Number of Cells per Compound",
    y = "Frequency (Number of Compounds)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

# Add vertical lines and print color legend to console
reference_lines <- c()

if(!is.na(dmso_avg)) {
  histogram_plot <- histogram_plot + 
    geom_vline(xintercept = dmso_avg, color = "green", linetype = "dotted", linewidth = 1)
  reference_lines <- c(reference_lines, "Green dotted line: DMSO")
}
if(!is.na(nocodazole_avg)) {
  histogram_plot <- histogram_plot + 
    geom_vline(xintercept = nocodazole_avg, color = "purple", linetype = "dotted", linewidth = 1)
  reference_lines <- c(reference_lines, "Purple dotted line: Nocodazole")
}
if(!is.na(tetrandrine_avg)) {
  histogram_plot <- histogram_plot + 
    geom_vline(xintercept = tetrandrine_avg, color = "red", linetype = "dotted", linewidth = 1)
  reference_lines <- c(reference_lines, "Red dotted line: Tetrandrine")
}

# Display the plot
print(histogram_plot)
```

*Insert analysis here*

## Cell Count Standard Deviation Distribution

```{r cell-count-sd-histogram}
# Create histogram of cell count standard deviation distribution
cat("\nCreating histogram of compound cell count standard deviation distribution...\n")

sd_plot <- ggplot(compound_counts, aes(x = cell_count_sd)) +
  geom_histogram(bins = 30, fill = "darkred", color = "black", alpha = 0.7) +
  labs(
    title = "Distribution of Cell Count Standard Deviation per Compound",
    x = "Standard Deviation of Cell Counts per Compound",
    y = "Frequency (Number of Compounds)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

# Add vertical lines for standard deviations and print color legend to console
sd_reference_lines <- c()

if(!is.na(dmso_sd)) {
  sd_plot <- sd_plot + 
    geom_vline(xintercept = dmso_sd, color = "green", linetype = "dotted", linewidth = 1)
  sd_reference_lines <- c(sd_reference_lines, "Green dotted line: DMSO")
}
if(!is.na(nocodazole_sd)) {
  sd_plot <- sd_plot + 
    geom_vline(xintercept = nocodazole_sd, color = "purple", linetype = "dotted", linewidth = 1)
  sd_reference_lines <- c(sd_reference_lines, "Purple dotted line: Nocodazole")
}
if(!is.na(tetrandrine_sd)) {
  sd_plot <- sd_plot + 
    geom_vline(xintercept = tetrandrine_sd, color = "red", linetype = "dotted", linewidth = 1)
  sd_reference_lines <- c(sd_reference_lines, "Red dotted line: Tetrandrine")
}

# Print combined reference line legend to console
if(length(reference_lines) > 0 || length(sd_reference_lines) > 0) {
  cat("\nReference lines on histograms:\n")
  cat("  Green dotted lines: DMSO\n")
  cat("  Purple dotted lines: Nocodazole\n")
  cat("  Red dotted lines: Tetrandrine\n")
}

# Display the standard deviation plot
print(sd_plot)
```

*Insert analysis here*

## Site-Specific Compound Analysis

```{r site-specific-analysis}
# Create compound analysis separated by site
if("Metadata_Site" %in% colnames(df)) {
  cat("\nCreating site-specific compound analysis...\n")
  
  compound_counts_by_site <- df %>%
    group_by(!!sym(compound_col), Metadata_Site) %>%
    summarise(
      cell_count = mean(Metadata_Object_Count, na.rm = TRUE),
      cell_count_sd = sd(Metadata_Object_Count, na.rm = TRUE),
      .groups = 'drop'
    )
}
```

*Insert analysis here*

## Site-Specific Histograms - Mean Cell Counts

```{r site-histogram-means}
# Create faceted histogram for mean cell counts by site
mean_by_site_plot <- ggplot(compound_counts_by_site, aes(x = cell_count)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
  facet_wrap(~ Metadata_Site, scales = "free_y", labeller = label_both) +
  labs(
    title = "Distribution of Mean Cell Counts per Compound by Site",
    x = "Average Number of Cells per Compound",
    y = "Frequency (Number of Compounds)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    strip.text = element_text(size = 11, face = "bold")
  )

# Calculate site-specific compound type averages
site_compound_averages <- df %>%
  filter(Metadata_Compound_type %in% c("DMSO", "Nocodazole", "Tetrandrine")) %>%
  group_by(Metadata_Site, Metadata_Compound_type) %>%
  summarise(
    avg_cell_count = mean(Metadata_Object_Count, na.rm = TRUE),
    .groups = 'drop'
  )

# Add vertical lines for site-specific means
mean_by_site_plot <- mean_by_site_plot + 
  geom_vline(data = site_compound_averages, 
             aes(xintercept = avg_cell_count, color = Metadata_Compound_type), 
             linetype = "dotted", linewidth = 1) +
  scale_color_manual(
    values = c("DMSO" = "green", "Nocodazole" = "purple", "Tetrandrine" = "red")
  ) +
  theme(legend.position = "none")

print(mean_by_site_plot)
```

*Insert analysis here*

## Site-Specific Histograms - Standard Deviations

```{r site-histogram-sds}
# Create faceted histogram for standard deviation by site
sd_by_site_plot <- ggplot(compound_counts_by_site, aes(x = cell_count_sd)) +
  geom_histogram(bins = 30, fill = "darkred", color = "black", alpha = 0.7) +
  facet_wrap(~ Metadata_Site, scales = "free_y", labeller = label_both) +
  labs(
    title = "Distribution of Cell Count Standard Deviation per Compound by Site",
    x = "Standard Deviation of Cell Counts per Compound",
    y = "Frequency (Number of Compounds)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    strip.text = element_text(size = 11, face = "bold")
  )

# Calculate site-specific compound type standard deviations
site_compound_sds <- df %>%
  filter(Metadata_Compound_type %in% c("DMSO", "Nocodazole", "Tetrandrine")) %>%
  group_by(Metadata_Site, Metadata_Compound_type) %>%
  summarise(
    sd_cell_count = sd(Metadata_Object_Count, na.rm = TRUE),
    .groups = 'drop'
  )

# Add vertical lines for site-specific SDs
sd_by_site_plot <- sd_by_site_plot + 
  geom_vline(data = site_compound_sds, 
             aes(xintercept = sd_cell_count, color = Metadata_Compound_type), 
             linetype = "dotted", linewidth = 1) +
  scale_color_manual(
    values = c("DMSO" = "green", "Nocodazole" = "purple", "Tetrandrine" = "red")
  ) +
  theme(legend.position = "none")

# Print combined reference line legend for site plots
cat("\nReference lines on site-specific histograms:\n")
cat("  Green dotted lines: DMSO (site-specific values)\n")
cat("  Purple dotted lines: Nocodazole (site-specific values)\n")
cat("  Red dotted lines: Tetrandrine (site-specific values)\n")

# Display the site plots
print(sd_by_site_plot)
```

*Insert analysis here*

## Well Position Grid Analysis

```{r well-position-setup}
# CREATE WELL POSITION GRID ANALYSIS
if("Metadata_Row" %in% colnames(df) && "Metadata_Col" %in% colnames(df)) {
  cat("\nCreating well position grid analysis...\n")
  
  # Create unique well identifier
  df_wells <- df %>%
    mutate(Well_ID = paste0(Metadata_Row, sprintf("%02d", as.numeric(Metadata_Col)))) %>%
    mutate(Well_Number = match(Well_ID, sort(unique(Well_ID))))
  
  # Calculate statistics per well
  well_stats <- df_wells %>%
    group_by(Metadata_Row, Metadata_Col, Well_ID, Well_Number) %>%
    summarise(
      avg_cell_count = mean(Metadata_Object_Count, na.rm = TRUE),
      sd_cell_count = sd(Metadata_Object_Count, na.rm = TRUE),
      observations = n(),
      .groups = 'drop'
    ) %>%
    arrange(Well_Number)
  
  # Display well mapping
  cat("\nFirst 20 well mappings:\n")
  print(head(well_stats[c("Well_ID", "Well_Number", "avg_cell_count")], 20))
  
  # Create 16x24 grid (assuming standard 384-well plate: A-P rows, 1-24 columns)
  # Convert row letters to numbers for plotting
  well_stats <- well_stats %>%
    mutate(
      Row_Num = match(Metadata_Row, LETTERS[1:16]),
      Col_Num = as.numeric(Metadata_Col)
    )
}
```

*Insert analysis here*

## Well Position Heatmaps by Site

```{r well-heatmaps-by-site}
# Create the heatmap separated by site
if("Metadata_Site" %in% colnames(df)) {
  cat("\nCreating 16x24 grid heatmap of average cell counts by site...\n")
  
  # Calculate statistics per well per site
  well_stats_by_site <- df %>%
    mutate(Well_ID = paste0(Metadata_Row, sprintf("%02d", as.numeric(Metadata_Col)))) %>%
    group_by(Metadata_Site, Metadata_Row, Metadata_Col, Well_ID) %>%
    summarise(
      avg_cell_count = mean(Metadata_Object_Count, na.rm = TRUE),
      sd_cell_count = sd(Metadata_Object_Count, na.rm = TRUE),
      observations = n(),
      .groups = 'drop'
    ) %>%
    mutate(
      Row_Num = match(Metadata_Row, LETTERS[1:16]),
      Col_Num = as.numeric(Metadata_Col)
    )
  
  well_heatmap <- ggplot(well_stats_by_site, aes(x = Col_Num, y = Row_Num, fill = avg_cell_count)) +
    geom_tile(color = "white", size = 0.1) +
    facet_wrap(~ Metadata_Site, labeller = label_both) +
    scale_fill_gradient(low = "blue", high = "red", name = "Average\nCell Count") +
    scale_x_continuous(breaks = seq(1, 24, by = 4), expand = c(0,0)) +
    scale_y_continuous(breaks = seq(1, 16, by = 2), labels = LETTERS[seq(1, 16, by = 2)], expand = c(0,0), trans = "reverse") +
    labs(
      title = "Average Cell Count by Well Position (16x24 Grid) by Site",
      x = "Column",
      y = "Row"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10),
      panel.grid = element_blank(),
      strip.text = element_text(size = 11, face = "bold")
    ) +
    coord_fixed()
  
  # Display the heatmap
  print(well_heatmap)
}
```

*Insert analysis here*

## Well Position Standard Deviation Heatmaps

```{r well-sd-heatmaps}
# Create heatmap for standard deviation by site
cat("\nCreating 16x24 grid heatmap of cell count standard deviation by site...\n")

well_sd_heatmap <- ggplot(well_stats_by_site, aes(x = Col_Num, y = Row_Num, fill = sd_cell_count)) +
  geom_tile(color = "white", size = 0.1) +
  facet_wrap(~ Metadata_Site, labeller = label_both) +
  scale_fill_gradient(low = "blue", high = "red", name = "Cell Count\nStd Dev") +
  scale_x_continuous(breaks = seq(1, 24, by = 4), expand = c(0,0)) +
  scale_y_continuous(breaks = seq(1, 16, by = 2), labels = LETTERS[seq(1, 16, by = 2)], expand = c(0,0), trans = "reverse") +
  labs(
    title = "Cell Count Standard Deviation by Well Position (16x24 Grid) by Site",
    x = "Column",
    y = "Row"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    panel.grid = element_blank(),
    strip.text = element_text(size = 11, face = "bold")
  ) +
  coord_fixed()

# Display the SD heatmap
print(well_sd_heatmap)
```

*Insert analysis here*

## Well Position Summary Statistics

```{r well-summary-stats}
# Print site-specific summary statistics
site_well_summary <- well_stats_by_site %>%
  group_by(Metadata_Site) %>%
  summarise(
    unique_wells = n(),
    min_cell_count = round(min(avg_cell_count, na.rm = TRUE), 2),
    max_cell_count = round(max(avg_cell_count, na.rm = TRUE), 2),
    mean_cell_count = round(mean(avg_cell_count, na.rm = TRUE), 2),
    min_sd = round(min(sd_cell_count, na.rm = TRUE), 2),
    max_sd = round(max(sd_cell_count, na.rm = TRUE), 2),
    mean_sd = round(mean(sd_cell_count, na.rm = TRUE), 2),
    .groups = 'drop'
  )

cat("\nWell position summary by site:\n")
print(site_well_summary)

# Print summary statistics
cat("\nWell position summary:\n")
cat("Total unique wells:", nrow(well_stats), "\n")
cat("Well number range:", min(well_stats$Well_Number), "to", max(well_stats$Well_Number), "\n")
cat("Average cell count range:", round(min(well_stats$avg_cell_count), 2), "to", round(max(well_stats$avg_cell_count), 2), "\n")
```

*Insert analysis here*

## 7 Plates Analysis - Raw Data

```{r seven-plates-setup}
# ------- 7 plates separately; raw, mean, sd, standardize by site ------     
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(ggplot2); library(stringr)
})

# =========================
# Config
# =========================
# ============================================================================
site_to_plot    <- "FMP"     # <-- change to the site you want
# ============================================================================
plates_per_site <- 7
reps_per_site   <- 4
rep_levels      <- paste0("R", 1:reps_per_site)

controls <- c("DMSO", "Nocodazole", "Tetrandrine")
control_color_map <- c("DMSO"="green","Nocodazole"="purple","Tetrandrine"="red",
                       "Other"="grey80","Missing"="black")
missing_fill_color <- "grey30"
tile_border_width  <- 0.25
```

```{r seven-plates-checks}
# =========================
# Checks
# =========================
stopifnot(exists("df"))
req_cols <- c("Metadata_Site","Metadata_Row","Metadata_Col",
              "Metadata_Object_Count","Metadata_Compound_type")
miss <- setdiff(req_cols, names(df))
if (length(miss)) stop("Missing cols: ", paste(miss, collapse=", "))

plate_col <- c("Metadata_Unique_plate","Metadata_Plate_num","Metadata_Plate",
               "Plate","PlateID","Barcode","Metadata_Barcode")
plate_col <- plate_col[plate_col %in% names(df)][1]
if (is.na(plate_col)) stop("Set `plate_col` to your plate ID column.")
```

```{r seven-plates-parsing}
# =========================
# Parse Plate & Replicate from plate string
# =========================
df_parsed <- df %>%
  mutate(
    Metadata_Row = as.character(Metadata_Row),
    Metadata_Col = suppressWarnings(as.numeric(Metadata_Col)),
    PlateRaw     = as.character(.data[[plate_col]]),
    .rep_digit   = str_match(PlateRaw, "(?i)(?:^|[_-])R\\s*([1-4])$")[,2],
    Replicate    = ifelse(!is.na(.rep_digit), paste0("R", str_trim(.rep_digit)), NA_character_),
    Plate        = ifelse(!is.na(Replicate),
                          str_replace(PlateRaw, "(?i)[_-]R\\s*[1-4]$", ""),
                          PlateRaw)
  ) %>%
  select(-.rep_digit) %>%
  filter(!is.na(Metadata_Col), Metadata_Col>=1, Metadata_Col<=24,
         Metadata_Row %in% LETTERS[1:16])

to_well_coords <- function(d) d %>% mutate(Row_Num = match(Metadata_Row, LETTERS[1:16]),
                                           Col_Num = Metadata_Col)

# Plate order helper (within THIS site)
plate_order_for_site <- function(site_name) {
  p <- df_parsed %>% filter(Metadata_Site==site_name) %>% distinct(Plate) %>% pull(Plate)
  if (!length(p)) return(character(0))
  num <- suppressWarnings(as.numeric(str_extract(p, "\\d+")))
  p[order(is.na(num), num, p)]
}
```

```{r seven-plates-build-grid}
# =========================
# Build complete **4 × 7** layout for ONE site
# =========================
build_site_grid <- function(site_name) {
  wells <- expand_grid(Metadata_Row=LETTERS[1:16], Metadata_Col=1:24)
  
  # pick exactly 7 plates (pad with placeholders if fewer exist)
  p_order <- plate_order_for_site(site_name)
  sel_plates <- head(p_order, plates_per_site)
  if (length(sel_plates) < plates_per_site) {
    sel_plates <- c(sel_plates,
                    paste0("MISSING_PLATE_", seq_len(plates_per_site - length(sel_plates))))
  }
  
  # skeleton: 7 plates × 4 reps × wells
  skel <- expand_grid(
    Metadata_Site = site_name,
    Plate         = sel_plates,
    Replicate     = rep_levels
  ) %>%
    crossing(wells) %>%
    to_well_coords() %>%
    mutate(
      PlateColIdx = match(Plate, sel_plates),          # columns = plates
      RepRowIdx   = match(Replicate, rep_levels),      # rows    = replicates
      PlateFacet  = factor(sprintf("Plate %d\n%s", PlateColIdx, Plate),
                           levels = sprintf("Plate %d\n%s", 1:plates_per_site, sel_plates)),
      RepFacet    = factor(sprintf("Rep %d\n%s", RepRowIdx, Replicate),
                           levels = sprintf("Rep %d\n%s", 1:reps_per_site, rep_levels))
    )
  
  # actual averages and dominant control (filtered to this site)
  dat <- df_parsed %>%
    filter(Metadata_Site==site_name) %>%
    group_by(Metadata_Site, Plate, Replicate, Metadata_Row, Metadata_Col) %>%
    summarise(avg_cell_count=mean(Metadata_Object_Count, na.rm=TRUE), .groups="drop")
  
  ctrl <- df_parsed %>%
    filter(Metadata_Site==site_name) %>%
    group_by(Metadata_Site, Plate, Replicate, Metadata_Row, Metadata_Col, Metadata_Compound_type) %>%
    summarise(n=dplyr::n(), .groups="drop") %>%
    group_by(Metadata_Site, Plate, Replicate, Metadata_Row, Metadata_Col) %>%
    slice_max(order_by=n, n=1, with_ties=FALSE) %>%
    ungroup() %>%
    mutate(control_label = ifelse(Metadata_Compound_type %in% controls,
                                  Metadata_Compound_type, "Other")) %>%
    select(Metadata_Site, Plate, Replicate, Metadata_Row, Metadata_Col, control_label)
  
  merged <- skel %>%
    left_join(dat,  by=c("Metadata_Site","Plate","Replicate","Metadata_Row","Metadata_Col")) %>%
    left_join(ctrl, by=c("Metadata_Site","Plate","Replicate","Metadata_Row","Metadata_Col"))
  
  if (!"control_label" %in% names(merged)) merged$control_label <- NA_character_
  
  merged %>%
    mutate(
      is_missing    = is.na(avg_cell_count),
      control_label = ifelse(is_missing, "Missing", control_label)
      # keep avg_cell_count = NA so fill uses `missing_fill_color`
    )
}

site_df <- build_site_grid(site_to_plot)
```

```{r seven-plates-raw-plot}
# =========================
# 1) RAW (single site) — **4 × 7** (rows = replicates, cols = plates)
# =========================
p_raw <- ggplot(site_df, aes(x=Col_Num, y=Row_Num)) +
  geom_tile(aes(fill=avg_cell_count, color=control_label), linewidth=tile_border_width) +
  facet_grid(RepFacet ~ PlateFacet, switch="y", drop=FALSE) +   # <<< 4 × 7 layout
  scale_fill_gradient(low="blue", high="red", name="Avg cell count",
                      na.value=missing_fill_color) +
  scale_color_manual(values=control_color_map, drop=FALSE, name="Border") +
  scale_x_continuous(breaks=seq(1,24,4), expand=c(0,0)) +
  scale_y_continuous(breaks=seq(1,16,2), labels=LETTERS[seq(1,16,2)],
                     trans="reverse", expand=c(0,0)) +
  labs(
    title=paste0("RAW — Site: ", site_to_plot, "  (4 Replicates × 7 Plates)"),
    subtitle="Missing plate–replicate wells are shaded in a distinct color; borders show control type",
    x="Column", y="Row"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title=element_text(hjust=0.5, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(face="bold")) +
  coord_fixed()
print(p_raw)
```

*Insert analysis here*

## 7 Plates Analysis - Average Across Replicates

```{r seven-plates-avg-plot}
# =========================
# 2) AVG across replicates (single site) — **1 × 7** (one row, 7 columns)
# =========================
avg_site <- site_df %>%
  group_by(Metadata_Site, Plate, PlateFacet, Metadata_Row, Metadata_Col, Row_Num, Col_Num) %>%
  summarise(avg_over_reps = mean(avg_cell_count, na.rm=TRUE), .groups="drop")

p_avg <- ggplot(avg_site, aes(x=Col_Num, y=Row_Num, fill=avg_over_reps)) +
  geom_tile(color="white", linewidth=tile_border_width) +
  facet_grid(. ~ PlateFacet, switch="y", drop=FALSE) +          # <<< 1 × 7
  scale_fill_gradient(low="blue", high="red", name="Avg over reps",
                      na.value=missing_fill_color) +
  scale_x_continuous(breaks=seq(1,24,4), expand=c(0,0)) +
  scale_y_continuous(breaks=seq(1,16,2), labels=LETTERS[seq(1,16,2)],
                     trans="reverse", expand=c(0,0)) +
  labs(
    title=paste0("AVG across replicates — Site: ", site_to_plot),
    subtitle="One row × 7 plates; grey indicates missing",
    x="Column", y="Row"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title=element_text(hjust=0.5, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(face="bold")) +
  coord_fixed()
print(p_avg)
```

*Insert analysis here*

## 7 Plates Analysis - Normalized Z-Scores

```{r seven-plates-normalized}
# =========================
# 3) NORMALIZED (per-site z-score of AVG) — **1 × 7**
# =========================
# -------- NORMALIZED (per-site z-score of AVG) — robust against SD == 0/NA -----
# avg_site is the 7×1 dataset built just above (per-plate, per-well averages)

# Helper: safe z-score on a numeric vector
safe_z <- function(x) {
  m <- mean(x, na.rm = TRUE)
  s <- sd(x, na.rm = TRUE)
  if (!is.finite(s) || s <= 0) {
    # fall back to mean-centering so we still see structure
    return(x - m)
  }
  (x - m) / s
}

# Compute z per site (single site selected upstream)
avg_norm <- avg_site %>%
  mutate(z_over_reps = safe_z(avg_over_reps))

# Optional: sanity printout
site_stats <- avg_site %>%
  summarise(
    n_non_na = sum(!is.na(avg_over_reps)),
    site_mean = mean(avg_over_reps, na.rm = TRUE),
    site_sd   = sd  (avg_over_reps, na.rm = TRUE)
  )
print(site_stats)
```

```{r seven-plates-normalized-plot}
# Plot (unchanged layout: 1 × 7)
p_norm <- ggplot(avg_norm, aes(x = Col_Num, y = Row_Num, fill = z_over_reps)) +
  geom_tile(color = "white", linewidth = tile_border_width) +
  facet_grid(. ~ PlateFacet, switch = "y", drop = FALSE) +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red", midpoint = 0,
    name = "Z-score (AVG)", na.value = missing_fill_color
  ) +
  scale_x_continuous(breaks = seq(1, 24, 4), expand = c(0, 0)) +
  scale_y_continuous(
    breaks = seq(1, 16, 2),
    labels = LETTERS[seq(1, 16, 2)],
    trans  = "reverse", expand = c(0, 0)
  ) +
  labs(
    title = paste0("NORMALIZED (Z-score of AVG) — Site: ", site_to_plot),
    subtitle = "Per-site z = (AVG − mean)/sd; falls back to mean-centering if sd ≤ 0 or NA",
    x = "Column", y = "Row"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold")
  ) +
  coord_fixed()

print(p_norm)
```

*Insert analysis here*

## Correlation of Cell Counts for well locations

```{r}
# =========================
# Plate × Plate (7×7) correlations by wells — per site
# Averaging 4 replicates first; Pearson r with pairwise complete wells
# =========================

# Sites to analyze (use all present; if you want just 4 specific sites, subset here)
sites <- sort(unique(df_parsed$Metadata_Site))

plots_corr <- list()

for (site in sites) {
  # 1) Pick exactly 7 plates for this site (same ordering logic you used)
  p_order    <- plate_order_for_site(site)
  sel_plates <- head(p_order, plates_per_site)
  if (length(sel_plates) < plates_per_site) {
    warning("Site ", site, " has only ", length(sel_plates),
            " plate(s); correlation will reflect available plates.")
  }

  # 2) Average across replicates per (Plate, Row, Col)
  avg_site <- df_parsed %>%
    dplyr::filter(Metadata_Site == site, Plate %in% sel_plates) %>%
    dplyr::group_by(Plate, Metadata_Row, Metadata_Col) %>%
    dplyr::summarise(avg_cell_count = mean(Metadata_Object_Count, na.rm = TRUE),
                     .groups = "drop") %>%
    dplyr::mutate(
      # stable well id and NA cleanup (mean of all-NA -> NaN -> NA)
      WellID = sprintf("%s%02d", Metadata_Row, Metadata_Col),
      avg_cell_count = dplyr::na_if(avg_cell_count, NaN)
    )

  # 3) Make a wells × plates table (rows = wells, cols = plates in your chosen order)
  mat_df <- avg_site %>%
    dplyr::select(WellID, Plate, avg_cell_count) %>%
    tidyr::pivot_wider(names_from = Plate, values_from = avg_cell_count) %>%
    dplyr::select(tidyselect::any_of(sel_plates))  # enforce order; ignore placeholders

  # Guard: need at least 2 plates to correlate
  if (ncol(mat_df) < 2) {
    message("Skipping site ", site, " (fewer than 2 plates after filtering).")
    next
  }

  # 4) Correlation across plates (using the well vectors)
  corr_mat <- stats::cor(mat_df, use = "pairwise.complete.obs", method = "pearson")

  # 5) Pretty axis labels consistent with your plate facets
  plate_labels <- sprintf("Plate %d\n%s", seq_along(colnames(corr_mat)), colnames(corr_mat))
  colnames(corr_mat) <- plate_labels
  rownames(corr_mat) <- plate_labels

  # 6) Tidy + plot
  corr_long <- as.data.frame(as.table(corr_mat)) %>%
    dplyr::rename(Plate_X = Var1, Plate_Y = Var2, corr = Freq)

  p <- ggplot(corr_long, aes(x = Plate_X, y = Plate_Y, fill = corr)) +
    geom_tile(color = "grey20", linewidth = 0.3) +
    geom_text(aes(label = sprintf("%.2f", corr)), size = 3) +
    scale_fill_gradient2(
      low = "blue", mid = "white", high = "red",
      midpoint = 0, limits = c(-1, 1), name = "Pearson r"
    ) +
    coord_equal() +
    labs(
      title = paste0("Plate × Plate Correlation by Wells — Site: ", site),
      subtitle = "Well-wise cell counts averaged over 4 replicates; pairwise-complete Pearson correlations",
      x = NULL, y = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      panel.grid  = element_blank(),
      plot.title  = element_text(hjust = 0.5, face = "bold")
    )

  print(p)
  plots_corr[[site]] <- p

  # 7) Console summary (optional)
  offdiag <- corr_mat[upper.tri(corr_mat) | lower.tri(corr_mat)]
  cat("\nSite:", site,
      "\n  Plates analyzed:", ncol(corr_mat),
      "\n  Mean off-diagonal r:", round(mean(offdiag, na.rm = TRUE), 3),
      "\n  Min r:", round(min(offdiag, na.rm = TRUE), 3),
      "  Max r:", round(max(offdiag, na.rm = TRUE), 3), "\n")
}

```

*Insert analysis here*

## Box-Plot of Combounds All Sites 
```{r Boxplots All Sites}
#normalize across-sites

df_compound_transform <- df %>% 
  mutate(Metadata_Compound_type = if_else(Metadata_Compound_type == "EOS", Metadata_EOS, Metadata_Compound_type))

df_norm <- df_compound_transform %>%
  group_by(Metadata_Site) %>%
  mutate(
    z_value = (Metadata_Object_Count - mean(Metadata_Object_Count, na.rm = TRUE)) / sd(Metadata_Object_Count, na.rm = TRUE)
  ) %>%
  ungroup()

# find top 5 compounds by median
top5_compounds <- df_norm %>%
  group_by(Metadata_Compound_type) %>%
  summarise(med_val = median(z_value, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(med_val)) %>%
  slice_head(n = 5) %>%
  pull(Metadata_Compound_type)

#filter data to those compounds and plot
df_norm %>%
  filter(Metadata_Compound_type %in% union(top5_compounds, c("Nocodazole", "Tetrandrine")) ) %>%
  ggplot(aes(x = Metadata_Compound_type, y = z_value)) +
  geom_boxplot() +
  labs(title = "Top 5 Compounds by Median Value")


# find top 5 compounds by median
bottom5_compounds <- df_norm %>%
  group_by(Metadata_Compound_type) %>%
  summarise(med_val = median(z_value, na.rm = TRUE), .groups = "drop") %>%
  arrange(med_val) %>%
  slice_head(n = 5) %>%
  pull(Metadata_Compound_type)

#filter data to those compounds and plot
df_norm %>%
  filter(Metadata_Compound_type %in% union(bottom5_compounds, c("DMSO")) ) %>%
  ggplot(aes(x = Metadata_Compound_type, y = z_value)) +
  geom_boxplot() +
  labs(title = "Top 5 Compounds by Median Value")


```

*Insert analysis here*
